---
title: "Blog Post 3"
description: |
  This blog post differs to the first two as it is a self-reflection focusing on a previously constructed shiny application, the purpose of which was to explore COVID-19 related data in the United States of America (USA). Rather than analysing the findings made possible by the app, a self-review of the analysis process will be undertaken focusing on four key questions.
preview: images/1140-american-flag-myths.jpg
author:
  - name: Lachlan Moody 27809951
    affiliation: Monash University
    affiliation_url: https://www.monash.edu/
date: 10-10-2020
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Shiny Application Introduction

The shiny application displayed below was produced for the purpose of assessment in ETC5523 and was made using the R package [Shiny]( https://shiny.rstudio.com/) which allows for the creation of interactive web apps. The application is currently hosted on [shinyapps.io](https://ltmoo2.shinyapps.io/shiny-assessment-ltmoo2/?_ga=2.89174936.2059005684.1602323177-246441122.1596523416).  

The application was designed to allow users to explore COVID-19 related data in the United States of America. Data relating to both the total number of cases and the total number of deaths attributed to the virus were incorporated for all the states in the country (excluding Hawaii and Alaska) and were displayed in three distinct visualisations for each measure. They were:  

1. A choropleth map displaying the spatial distribution of either total cases or total deaths as at the 28th of September 2020.  
2. A timeseries plot displaying the change in either total cases or total deaths over time at the state level between the 22nd of January and the 28th of September 2020, compared to the national average at the time.  
3. A table recording the data behind the timeseries plot displaying the readings for each day over the period for total cases and total deaths respectively compared to the national average.  

In addition to this, the power of the Shiny package is in its ability to make interactive visualisations that respond to user feedback. Three user controls were implemented into the application that control the above outputs. First a radio button was added to allow users to switch the measure displayed by all the graphics between ‘Total Cases’ and ‘Total Deaths’. Second, a dropdown list was added that allowed different states to be selected and compared in the timeseries and table output. Additionally, this input could be updated by clicking on the corresponding state on the map. Finally, a date slider was added that allowed the data displayed in the timeseries and table to be filtered to a specific date range.  

More detailed explorations are provided within the app itself should you wish to revisit it or explore it further before examining the accompanying review.

```{r, layout="l-screen-inset shaded"}
knitr::include_app('https://ltmoo2.shinyapps.io/shiny-assessment-ltmoo2/', height = '1000px')
```

## Self Review

However with this, and indeed any educational undertaking, self-reflection is an important part of the learning process. Holding this viewpoint in mind, the following four review questions were considered in the interest of expanding my understanding of the project itself and my learning processes as a whole.

### 1. What did you do well?



### 2. What are the points you can improve on?



### 3. What are some things that you’d like to add that’s plausible?



### 4. What are some things that you’d like to add that cannot be done yet due to limitation in data or technology?





