---
title: "Blog Post 3"
description: |
  This blog post differs to the first two as it is a self-reflection focusing on a previously constructed shiny application, the purpose of which was to explore COVID-19 related data in the United States of America (USA). Rather than analysing the findings made possible by the app, a self-review of the analysis process will be undertaken focusing on four key questions.
preview: images/1140-american-flag-myths.jpg
author:
  - name: Lachlan Moody 27809951
    affiliation: Monash University
    affiliation_url: https://www.monash.edu/
date: 10-10-2020
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Shiny Application Introduction

The shiny application displayed below was produced for the purpose of assessment in ETC5523 and was made using the R package [Shiny]( https://shiny.rstudio.com/) which allows for the creation of interactive web apps. The application is currently hosted on [shinyapps.io](https://ltmoo2.shinyapps.io/shiny-assessment-ltmoo2/?_ga=2.89174936.2059005684.1602323177-246441122.1596523416).  

The application was designed to allow users to explore COVID-19 related data in the United States of America. Data relating to both the total number of cases and the total number of deaths attributed to the virus were incorporated for all the states in the country (excluding Hawaii and Alaska) and were displayed in three distinct visualisations for each measure. They were:  

1. A choropleth map displaying the spatial distribution of either total cases or total deaths as at the 28th of September 2020.  
2. A timeseries plot displaying the change in either total cases or total deaths over time at the state level between the 22nd of January and the 28th of September 2020, compared to the national average at the time.  
3. A table recording the data behind the timeseries plot displaying the readings for each day over the period for total cases and total deaths respectively compared to the national average.  

In addition to this, the power of the Shiny package is in its ability to make interactive visualisations that respond to user feedback. Three user controls were implemented into the application that control the above outputs. First a radio button was added to allow users to switch the measure displayed by all the graphics between ‘Total Cases’ and ‘Total Deaths’. Second, a dropdown list was added that allowed different states to be selected and compared in the timeseries and table output. Additionally, this input could be updated by clicking on the corresponding state on the map. Finally, a date slider was added that allowed the data displayed in the timeseries and table to be filtered to a specific date range.  

More detailed explorations are provided within the app itself should you wish to revisit it or explore it further before examining the accompanying review.

```{r, layout="l-screen-inset shaded"}
knitr::include_app('https://ltmoo2.shinyapps.io/shiny-assessment-ltmoo2/', height = '1000px')
```

## Self Review

However with this, and indeed any educational undertaking, self-reflection is an important part of the learning process. Holding this viewpoint in mind, the following four review questions were considered in the interest of expanding my understanding of the project itself and my learning process as a whole.

### 1. What did you do well?



### 2. What are the points you can improve on?



### 3. What are some things that you’d like to add that’s plausible?
Taking the above points into consideration, there are several possible additions that could be made to the application in order to improve its functionality and use for exploring COVID-19 data for the purpose of analysis. One possible option is to expand the scope of the app to incorporate more, or even all, countries around the globe to provide comparisons internationally. Also, additional measures relating to the pandemic itself and different government’s responses to it, such as the number of tests or lockdown measures, could be added. This would allow for a greater number of questions to be addressed within the app.  

Alternatively, a more detailed view of the USA could have been provided, further breaking down the country into different regions or counties within states. This would allow for a more nuanced interpretation of the viruses spread as a distinction could be made between metropolitan and regional areas, leading to more accurate depictions of  virus clusters throughout the country.  

Additionally, going down either of these paths, a temporal component could be included in the spatial visualisation of the data which was not present in the original design. This would allow users to identify how the virus spread throughout the country, or indeed, throughout the world. This could be implemented either via an additional shiny input or through an animation that would show the progression from day-to-day. This would add a further layer of complexity and additional information to the produced graphic.  

### 4. What are some things that you’d like to add that cannot be done yet due to limitation in data or technology?
In order to gain a deeper understanding of COVID-19 and how it has spread throughout the USA, data containing information on the occurrence of individual cases and who it was transmitted to would be required. This would allow for a more detailed visualisation depicting exactly how the virus was able to radiate across the country in such a short period of time and infect such a large number of people. This would provide immense value as it would provide the exact path such viruses can take, thus allowing for plans and safeguards to be put in place for the future. However, there are several reasons why such data is unavailable. Firstly, there are confidentiality and data security issues relating to such data as it would involve disclosing peoples medical records and personal information which may be used to identify them. This may lead to negative consequences for such individuals, especially if they were responsible for spreading COVID-19. Secondly, with [over 7,500,000 cases in the country] (https://covid.cdc.gov/covid-data-tracker/#cases_casesinlast7days) the technology required to accurately conduct contact tracing is likely to be near impossible with two many unconfirmed transmissions to accurately track the virus. 




