---
title: "Blog Post 4"
description: |
  This blog post expands on the theme of reflection and review from the third blog post. While that post focused on a self-assessment, this one will expand into the area of peer review, the context of which will be the previously completed R package assessment for ETC5523.
preview: images/covid19.jpg
author:
  - name: Lachlan Moody 27809951
    affiliation: Monash University
    affiliation_url: https://www.monash.edu/
date: 10-25-2020
output:
  distill::distill_article:
    self_contained: false
bibliography: biblio.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Package Overview & Author Details
The purpose of this blog post is to provide a peer review of the R package produced by fellow-Monash student Chengzhi Ye, titled **`CovidYe`**, contained within the GitHub repository [etc5523-2020/r-package-assessment-Year-97](https://github.com/etc5523-2020/r-package-assessment-Year-97).   

This package was constructed in accordance with the instructions for assignment 3 for ETC5523, and its primary function is to provide a shiny application that allows users to explore COVID-19 related data in a package format. More specifically, according to the readme file, the package provides information on the countries most affected by the virus and contains one data set and three additional functions designed to create visualisations with interactive elements presented using a shiny application.  

For the purposes of this review, the package was installed into R Studio on the 25/10/2020 using the following command provided in the readme:

```{r, eval=FALSE, echo=TRUE}
# install.packages("devtools")
devtools::install_github("etc5523-2020/r-package-assessment-Year-97")
```

## Package Execution



## Package Improvements
There were a number of issues that were found when trying to navigate through the package and launch the shiny application itself that if fixed would improve the functionality and usability of the **`CovidYe`** package.  
Firstly, in both the readme file on GitHub and [vignette website]( https://etc5523-2020.github.io/r-package-assessment-Year-97/articles/COVID.html) there is no instruction to load the library in before being told to launch the app. While it may be a reasonable assumption that only experienced R users would be downloading this package, an extra section containing the following code below would expand the accessibility of what has been created for all users of the software.

```{r, eval=FALSE, echo=TRUE}
library(CovidYe)
launch_app()
```

Secondly, and perhaps more of an important issue, the `launch_app()` function that had been created did not work initially when ran in a R console. Instead the error message ` Error in library(coronavirus) : there is no package called ‘coronavirus’` appeared. While I was able to get around this by installing this package separately, users unfamiliar with R may have faced difficulties getting the shiny application to run. This error was caused as `library(coronavirus` was called within the app.R file without being imported by the package. To resolve this issue the package `coronavirus`  could’ve been added to the package dependencies using the following code below:

```{r, eval=FALSE, echo=TRUE}
usethis::use_package("coronavirus")
```


An additional error was encountered when trying to run the `launch_app()` function after the aforementioned problem was resolved. The user-created function `Inputselect()` was causing the message `Error in shiny::selectInput(inputId = ID, "Which country do you choose",  :  argument "chosen_column" is missing, with no default` to appear in the console. Unfortunately this problem could not be resolved without downloading the [entire package from GitHub](https://github.com/etc5523-2020/r-package-assessment-Year-97) and altering the code within the app.R file to provide a default `”chosen_column”`. Luckily the documentation provided for this function made this an easy fix as the different acceptable values were clearly stated and were able to be included in the function call. Once done this allowed the application to be launched without error.  

Finally, regarding the app itself, while many detailed insights and visually impressive graphics are provided in a logical and coherent layout, it could have been improved with some additional stylistic choices. This could have been achieved by adding some simple CSS or even simply by using the shinythemes package[@themes].



## Key Learnings


## Review Checklist {.appendix}
**Documentation**  
The package includes all the following forms of documentation:  
- [ ] **Installation instructions:** for the package is found in README  
- [ ] **Vignette(s)** demonstrating major functionality that runs successfully locally  
- [ ] **Function Documentation:** for all exported functions in R help  
- [ ] **Examples** for all exported functions in R Help that run successfully locally  
**Functionality**  
- [ ] **Installation:** Installation succeeds as documented.  
- [ ] **Functionality:** Any functional claims of the software been confirmed.  
- [ ] **Automated tests:** Unit tests cover essential functions of the package
and a reasonable range of inputs and conditions. All tests pass on the local machine.