---
title: "Blog Post 4"
description: |
  This blog post expands on the theme of reflection and review from the third blog post. While that post focused on a self-assessment, this one will expand into the area of peer review, the context of which will be the previously completed R package assessment for ETC5523.
preview: images/covid19.jpg
author:
  - name: Lachlan Moody 27809951
    affiliation: Monash University
    affiliation_url: https://www.monash.edu/
date: 10-25-2020
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Package Overview & Author Details
The purpose of this blog post is to provide a peer review of the R package produced by fellow-Monash student Chengzhi Ye, titled **`CovidYe`**, contained within the GitHub repository [etc5523-2020/r-package-assessment-Year-97](https://github.com/etc5523-2020/r-package-assessment-Year-97).   

This package was constructed in accordance with the instructions for assignment 3 for ETC5523, and its primary function is to provide a shiny application that allows users to explore COVID-19 related data in a package format. More specifically, according to the readme file, the package provides information on the countries most affected by the virus and contains one data set and three additional functions designed to create visualisations with interactive elements presented using a shiny application.  

For the purposes of this review, the package was installed into R Studio on the 25/10/2020 using the following command provided in the readme:

```{r, eval=FALSE, echo=TRUE}
# install.packages("devtools")
devtools::install_github("etc5523-2020/r-package-assessment-Year-97")
```


