---
title: "Blog Post 4"
description: |
  This blog post expands on the theme of reflection and review from the third blog post. While that post focused on a self-assessment, this one will expand into the area of peer review, the context of which will be the previously completed R package assessment for ETC5523.
preview: images/covid19.jpg
author:
  - name: Lachlan Moody 27809951
    affiliation: Monash University
    affiliation_url: https://www.monash.edu/
date: 10-25-2020
output:
  distill::distill_article:
    self_contained: false
bibliography: biblio.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Package Overview & Author Details
The purpose of this blog post is to provide a peer review of the R package produced by fellow-Monash student Chengzhi Ye, titled **`CovidYe`**, contained within the GitHub repository [etc5523-2020/r-package-assessment-Year-97](https://github.com/etc5523-2020/r-package-assessment-Year-97).   

This package was constructed in accordance with the instructions for assignment 3 for ETC5523, and its primary function is to provide a shiny application that allows users to explore COVID-19 related data in a package format. More specifically, according to the readme file, the package provides information on the countries most affected by the virus and contains one data set and three additional functions designed to create visualisations with interactive elements presented using a shiny application.  

For the purposes of this review, the package was installed into R Studio on the 25/10/2020 using the following command provided in the readme:

```{r, eval=FALSE, echo=TRUE}
# install.packages("devtools")
devtools::install_github("etc5523-2020/r-package-assessment-Year-97")
```

## Package Execution



## Package Improvements
There were a number of issues that were found when trying to navigate through the package and launch the shiny application itself that if fixed would improve the functionality and usability of the **`CovidYe`** package.  
Firstly, in both the readme file on GitHub and [vignette website]( https://etc5523-2020.github.io/r-package-assessment-Year-97/articles/COVID.html) there is no instruction to load the library in before being told to launch the app. While it may be a reasonable assumption that only experienced R users would be downloading this package, an extra section containing the following code below would expand the accessibility of what has been created for all users of the software.

```{r, eval=FALSE, echo=TRUE}
library(CovidYe)
launch_app()
```

Secondly, and perhaps more of an important issue, the `launch_app()` function that had been created did not work initially when ran in a R console. Instead the error message ` Error in library(coronavirus) : there is no package called ‘coronavirus’` appeared. While I was able to get around this by installing this package separately, users unfamiliar with R may have faced difficulties getting the shiny application to run. This error was caused as `library(coronavirus` was called within the app.R file without being imported by the package. To resolve this issue the package `coronavirus`  could’ve been added to the package dependencies using the following code below:

```{r, eval=FALSE, echo=TRUE}
usethis::use_package("coronavirus")
```


An additional error was encountered when trying to run the `launch_app()` function after the aforementioned problem was resolved. The user-created function `Inputselect()` was causing the message `Error in shiny::selectInput(inputId = ID, "Which country do you choose",  :  argument "chosen_column" is missing, with no default` to appear in the console. Unfortunately this problem could not be resolved without downloading the [entire package from GitHub](https://github.com/etc5523-2020/r-package-assessment-Year-97) and altering the code within the app.R file to provide a default `”chosen_column”`. Luckily the documentation provided for this function made this an easy fix as the different acceptable values were clearly stated and were able to be included in the function call. Once done this allowed the application to be launched without error.  

Finally, regarding the app itself, while many detailed insights and visually impressive graphics are provided in a logical and coherent layout, it could have been improved with some additional stylistic choices. This could have been achieved by adding some simple CSS or even simply by using the shinythemes package[@themes].



## Key Learnings
There were several key learnings gained from conducting this review that I will take forward into my own package development and coding practices in the future.  
Firstly, there is a great need to be explicit in your instructions and comments when undertaking any programming activity. This is especially important in something such as a R package which is likely to be shared with other users. What may be clear to one person at a point in time may be confusing to another who was not involved in the development phase. This was evident in the discussion above regarding the missing `library()` call in the documentation which shows what can happen with poor instructions. On the other hand, the `Inputselect` issue was able to be resolved due to good documentation and clear instructions.  

Secondly, and somewhat combined with the above, is the need to understand your audience for what you are creating. While the issues discussed above are easily fixed by someone with knowledge of R and package development, some users may get stuck and never end up viewing what has been created. In comparison the app itself is well designed to cater to any audience as there are many detailed explanations and instructions provided throughout with easily interpretable plots.  

Finally, the importance of error checking and testing cannot be understated. What may work on the creators computer and software setup may fail on others. The issues faced could’ve been avoided by first following the created instructions on a fresh device or R session to see if the package could be successfully and easily run. This would’ve unveiled the missing dependency for the `coronavirus` package and the failure of the created `Inputselect` function thus creating a more user-friendly package.


## Review Checklist {.appendix}
**Documentation**  
The package includes all the following forms of documentation:  
- [x] **Installation instructions for the package is found in README:** Clear and easy to follow instructions are available in GitHub readme file.     
- [x] **Vignette(s) demonstrating major functionality that runs successfully locally:** Vignette provided online using `pkgdown` and contain information on the entire package and its various functions.  
- [x] **Function Documentation for all exported functions in R help:** All functions exported by the package have some form of documentation.  
- [ ] **Examples for all exported functions in R Help that run successfully locally:** Examples for function `Easykable` and `coronavirus` not found  

**Functionality**  
- [x] **Installation Installation succeeds as documented:** Package is able to be installed into R without error or issue.  
- [ ] **Functionality Any functional claims of the software been confirmed:** Function `launch_app` does not work on instal and `Inputselect` fails within app  
- [x] **Automated tests Unit tests cover essential functions of the package and a reasonable range of inputs and conditions. All tests pass on the local machine:** At least three tests for each user created function that cleverly test the validity of the outputs created and all pass without error.