---
title: "Blog Post 1"
description: |
  This blog post explores the impact of COVID-19 in Argentina. In particular, the various protective health measures introduced by the government will be evaluated based on their ability to effectively reduce people's movements and ultimately stop the spread of the virus.
preview: images/argentina-flag.jpg
author:
  - name: Lachlan Moody 27809951
    affiliation: Monash University
    affiliation_url: https://www.monash.edu/
date: 08-27-2020
output:
  distill::distill_article:
    self_contained: false
bibliography: biblio.bib
---
COVID-19 has been the topic of much discussion and interest over the course of 2020. Argentina has been no different with many neighboring South American countries suffering greatly at the hands of the virus. This blog post will explore the current situation in the country, and in particular how the national government has tackled this global pandemic.\  

While some reports have labeled their public health measures as “harsh” during a time where the country was already in economic trouble[@costabel_2020], ultimately its effect on reducing the spread of the virus requires further evaluation to weigh up the benefits of such an approach.


## Data Description
```{r setup, include=FALSE}
library(devtools)
#install_github("joachim-gassen/tidycovid19")
library(tidyverse)
library(tidycovid19)
library(zoo)
library(ggplot2)
library(kableExtra)
library(DT)
library(broom)
library(sparkline)



knitr::opts_chunk$set(echo = FALSE, message = FALSE)
sparkline(0)
```

To examine this subject in depth, the *tidycovid19* package [@covid] has been downloaded directly into R using the *devtools* package[@dev].\

The raw data files are provided in the GitHub repository joachim-gassen/tidycovid19[@git], and contains information collected from various sources such as Johns Hopkins University (JHU), the European Centre for Disease Prevention and Control, Apple Maps & Google Community Mobility Reports.\  

Observations are recorded at the country level for each day from the 31st of December 2019 up until the current day (the package is updates as information is released). In total, the downloaded dataset has 35 different variables each pertaining to different aspects of the virus, however for the proceeding analysis the following variables have been selected:\  

- country: name of country\
- date: calendar date\
- confirmed: confirmed Covid-19 cases as per JHU\
- deaths: confirmed Covid-19 deaths as per JHU\
- recovered: confirmed Covid-19 recoveries as per JHU\
- soc_dist: net number of social distancing measures\
- mov_rest: net number of movement restrictions\
- pub_health: net number of public health measures\
- gov_soc_econ: net number of social and economic measures\
- lockdown: net number of lockdown measures\
- apple_mtr_driving: Apple Maps usage for driving directions, as percentage relative to the baseline of Jan 13, 2020\
- apple_mtr_walking: Apple Maps usage for walking directions, as percentage relative to the baseline of Jan 13, 2020\
- gcmr_grocery_pharmacy: Google Community Mobility Reports data for the frequency that people visit grocery stores and pharmacies expressed as a percentage change relative to the baseline period Jan 3 - Feb 6, 2020\
- gcmr_parks: Google Community Mobility Reports data for the frequency that people visit parks expressed as a percentage change relative to the baseline period Jan 3 - Feb 6, 2020\
- gcmr_residential: Google Community Mobility Reports data for the frequency that people visit residential places expressed as a percentage change relative to the baseline period Jan 3 - Feb 6, 2020\
- gcmr_retail_recreation: Google Community Mobility Reports data for the frequency that people visit retail and recreation places expressed as a percentage change relative to the baseline period Jan 3 - Feb 6, 2020\
- gcmr_transit_stations: Google Community Mobility Reports data for the frequency that people visit transit stations expressed as a percentage change relative to the baseline period Jan 3 - Feb 6, 2020\
- gcmr_workplaces: Google Community Mobility Reports data for the frequency that people visit workplaces expressed as a percentage change relative to the baseline period Jan 3 - Feb 6, 2020 \

Once read into R, the Apple Maps usage data were converted from a relative percentage to a relative percentage change to match the Google data and a proportion of the country’s population infected and killed by the virus was calculated. Additionally, daily counts were constructed for each of the ‘confirmed’, ‘deaths’ and ‘recovered’ variables. The new variables were produced and added to the provided data using the *dplyr* package[@dplyr].


## Data Story
```{r data, include = FALSE}
arg_data <- download_merged_data(silent = TRUE, cached = TRUE) %>%
  filter(country == "Argentina")
  

arg_data_clean <- arg_data %>%
  mutate(lag_confirmed = lag(confirmed),
         daily_confirmed = confirmed - lag_confirmed,
         lag_deaths = lag(deaths),
         daily_deaths = deaths - lag_deaths,
         lag_recovered = lag(recovered),
         daily_recovered = recovered - lag_recovered,
         proportion_infected = confirmed/population,
         proportion_killed = deaths/population)
```

Before an evaluation of the various government initiatives can be, it is first prudent to explore the impact of the virus itself on the country.\

The table below provides an opportunity to explore some of the statistics provided in the dataset.\ 

Beginning with the first column, it can be seen that the period of time covered is from the 22nd of January through to the 29th of August. This provides quite a substantial amount of data from which to make inferences.\

The table, being interactive, is able to be sorted and filtered across any column. Beginning on the left hand side, it appears that at its peak Argentina was recording 11,717 cases per day. Perhaps worryingly, all of the largest daily increases have been recorded at the tail end of the data set(as indicated by the row number), indicating that the virus is still not under control in the country.\

Moving along to the next column, sorting by 'Total Confirmed Cases', at the last recorded date, Argentina had over 400,000 confirmed infections. While a large number, when comparing it to the 'Proportion Infected' further to the right, it is notable that this is still only 0.89% of the population.\

Further down the table to the right, it also can be seen that the most number of deaths recorded for a single day was 381 and the total amount of deaths is 8,353, when sorting by 'Daily Deaths' and 'Total Deaths' respectively. Similarly to the daily case numbers, many of the higher daily death counts are during the more recent dates in the data set.\

```{r interactive1}
arg_interactive_data <- arg_data_clean %>%
  mutate(apple_mtr_driving = apple_mtr_driving - 100,
         apple_mtr_walking = apple_mtr_walking - 100,
         proportion_infected = paste(round(100*proportion_infected, digits = 2),"%"),
         proportion_killed = paste(round(100*proportion_killed, digits = 2),"%")) %>%
  filter(date < "2020-08-30")

explore <- arg_interactive_data %>%
  select(Date = date,
         "Daily Confirmed Cases" = daily_confirmed,
         "Total Confirmed Cases" = confirmed,
         "Daily Recoveries" = daily_recovered,
         "Total Recoveries" = recovered,
         "Daily Deaths" = daily_deaths,
         "Total Deaths" = deaths,
         "Proportion Infected" = proportion_infected,
         "Proportion Killed" = proportion_killed)

datatable(explore, 
          escape = FALSE,
          filter = 'top',
          caption = htmltools::tags$caption( style = 'caption-side: top; text-align: center; color:black; font-size:200% ;','COVID19 case statistics in Argentina'),
          options = list(columnDefs = list(list(className = 'dt-right', targets = 2:9)),
                         fnDrawCallback = htmlwidgets::JS(
  '
function(){
  HTMLWidgets.staticRender();
}
'
)
       )) %>%
  spk_add_deps() 
```
\
Now that an understanding of the current situation has been gathered, the next stage is to analyse how the movements of Argentina's population has changed during this time. This will be accomplished using the Apple Maps data and Google Community Mobility Reports provided by the data package.\

Argentina's government acted much faster than its neighbors when news of the disease was still emerging, implementing "one of the strictest quarantines" in the region with only 128 active cases  which resulted in a total economic shutdown of the country[@costabel_2020].\

This impact is quite apparent in the 'Trend' section of the table below, which measures the change in traffic over time, with all measures recording a steep drop off in recorded traffic except for residential which saw the inverse. This reflects the situation in the country as most people were confined to their homes. As restrictions begun to ease over time, the trend lines are slowly returning to normal levels.\

Sorting the table by 'Average Change', it appears that the worst affected area was visitation to parks with most measures seeing a more than 40% decrease in traffic. Residential traffic on the other hand was the only one to experience an increase (this pattern is reflected by the graphs). Perhaps unsurprisingly, excluding residential, grocery and pharmacy traffic was the least effected. This is most likely because, similarly to Australia, essential services, such as pharmacies and drugstores, were some of the only business allowed to be open[@costabel_2020].

```{r inteactive2}
spk_test <- data.frame(
  Measure = c("driving","walking", "retail/recreation", "grocery/pharmacy", "parks", "transit/stations", "workplaces", "residential"),
  Trend = c(spk_chr(arg_interactive_data$apple_mtr_driving),
  spk_chr(arg_interactive_data$apple_mtr_walking),
  spk_chr(arg_interactive_data$gcmr_retail_recreation),
  spk_chr(arg_interactive_data$gcmr_grocery_pharmacy),
  spk_chr(arg_interactive_data$gcmr_parks),
  spk_chr(arg_interactive_data$gcmr_transit_stations),
  spk_chr(arg_interactive_data$gcmr_workplaces),
  spk_chr(arg_interactive_data$gcmr_residential)
))

spk_test  <- spk_test %>%
  arrange(Measure)

test2 <- arg_interactive_data %>%
  select(driving = apple_mtr_driving,
         walking = apple_mtr_walking,
         "grocery/pharmacy" = gcmr_grocery_pharmacy,
         parks = gcmr_parks,
         residential = gcmr_residential,
         "retail/recreation" = gcmr_retail_recreation,
         "transit/stations" = gcmr_transit_stations,
         "workplaces" = gcmr_workplaces,
         date) %>%
  pivot_longer(cols = driving:workplaces, names_to = "Measure") %>%
  filter(!is.na(value)) %>%
  group_by(Measure) %>%
  summarise("First Observation" = min(date),
            Observations = n(),
            "Average Change" = formatC(round(mean(value), digits =2), format = 'f', digits =2),
            "Standard Deviation" = formatC(round(sd(value), digits =2), format = 'f', digits =2))


test2 <- test2 %>%
  mutate("Average Change" = paste(`Average Change`,"%"),
         "Standard Deviation" = paste(`Standard Deviation`, "%"))

joint <- left_join(test2, spk_test)

datatable(joint, 
          escape = FALSE,
          filter = 'top',
          caption = htmltools::tags$caption( style = 'caption-side: top; text-align: center; color:black; font-size:200% ;','Effect of COVID 19 on public movement'),
          options = list(paging = FALSE, 
                         columnDefs = list(list(className = 'dt-right', targets = 2:5)),
                         fnDrawCallback = htmlwidgets::JS(
  '
function(){
  HTMLWidgets.staticRender();
}
'
)
       )) %>%
  spk_add_deps() 
```
So while it appears that the government has effectively reduced people's movements throughout the country, the real question is if it has had any effect on the spread of the virus.\

According to the Australian Government (and many prevailing health bodies around the world), COVID19 spreads in one of three ways; either through close contact with an infectious person, contact with water droplets from an infected person or touching objects that have droplets from an infected person[@aus].\

Thus the prevailing thought is that limiting people to essential contact only will limit the likelihood of them contacting the virus. The data provided divides such measures into 5 categories; social distancing measures, movement restrictions, public health measures, social/economic measures and lockdowns, providing current number of each in place. These measures have been used to construct 5 separate linear models, using the counts as an explanatory variable for the number of cases. The table below provides the resulting model statistics.


```{r static}
lm_soc_dist <-  lm(daily_confirmed ~ soc_dist, data = arg_data_clean)
lm_mov_rest <-  lm(daily_confirmed ~ mov_rest, data = arg_data_clean)
lm_pub_health <-  lm(daily_confirmed ~ pub_health, data = arg_data_clean)
lm_gov_soc_econ <-  lm(daily_confirmed ~ gov_soc_econ, data = arg_data_clean)
lm_lockdown <- lm(daily_confirmed ~ lockdown, data = arg_data_clean)

lm_tidy <- bind_rows(tidy(lm_soc_dist),
          tidy(lm_mov_rest),
          tidy(lm_pub_health),
          tidy(lm_gov_soc_econ),
          tidy(lm_lockdown)) %>%
  filter(term != "(Intercept)")


lm_glance <- bind_rows(glance(lm_soc_dist),
          glance(lm_mov_rest),
          glance(lm_pub_health),
          glance(lm_gov_soc_econ),
          glance(lm_lockdown))



lm_table <- bind_cols(lm_tidy, 
          lm_glance) %>%
  select(term, estimate, r.squared, "p value" = p.value...5)


lm_table_final <- lm_table %>%
  mutate(estimate = round(lm_table$estimate, digits = 2),
         r.squared = formatC(round(lm_table$r.squared, digits = 2), format = 'f', digits = 2),
         "p value" = formatC(round(lm_table$`p value`, digits = 2), format = 'f', digits =2)) %>%
  mutate(estimate = cell_spec(estimate, 
                              color = if_else(estimate < 0, "white", "black"), 
                              background = ifelse(estimate < 0, "green", "white")),
         r.squared = cell_spec(r.squared, color = ifelse(r.squared > 0.3, "white", "black"), 
                               background = ifelse(r.squared > 0.3, "green", "white")),
         "p value" = cell_spec(`p value`, color = ifelse(`p value` > 0.05, "black", "white"),
                               background = ifelse(lm_table$`p value` < 0.05, "green", "white"))) %>%
  kable(escape = F,
        caption = "Model comparisons for protective health measures", 
        format.args = list(digits = 2),
        align = "lrrr") %>%
  kable_styling(bootstrap_options = c("bordered","striped", "hover"))

lm_table_final


```
There are several key observations that can be made from this table. Firstly, the estimate tells us how much of an effect each one unit increase in the measure had on the daily number of cases. As the goal is to reduce this number, lower numbers are better. Taking this column in isolation, it would social distancing had the only impact on actually reducing the number of daily cases. However, when moving to the second column, it can be seen that the r squared value is extremely low, indicating that this model does not have much explanatory power. Furthermore, the p value is above the required significance level of 0.05 indicating that there is not enough evidence to state that this measure has an effect.\

So are none of the measures effective then?\

Perhaps not, because there is an important factor to consider. As the number of daily cases increases, naturally the government will respond by instituting more measures. So with that in mind, it may be that social/economic measures and movement restrictions are the most effective measures.\

Beginning with social/economic measures, it has the lowest of the remaining estimates and a p value below 0.05, meaning the measure had a significant effect. While the r squared is one of the lower values at 0.36, this is still quite good for data off this scale. This indicates that 36% of the change in daily cases can be explained by the number of social/economic measures in place. This is quite an interesting find, as while not a direct restriction on movement, it perhaps is effective as it provides an incentive or ability for people to remain home. This however is out of the scope of this blog post.\

Finally, movement restrictions had the next lower estimate and also a p value below the 0.05 cutoff. This is perhaps more expected as the key to stopping the virus spreading is to stop movement. This is reflected in a r squared value more than double that of social/economic measures at 0.68. 
